---
title: "Computer Support Practical Tasks"
date: "10/14/2024"
format:
  html:
    embed-resources: true
editor: source
---

## Introduction

For performing practical tasks **R** and **R Studio** have to be installed.

Links for downloading programs: 

1. R: <https://cran.r-project.org/>

2. R Studio: <https://posit.co/download/rstudio-desktop/>

The results of laboratory work must be placed in your repository on Github in the files `TaskX.qmd` (or `LabX.Rmd` if you prefer to use R markdown) and generated from it `TaskX.html`, where X is the number of the laboratory in the form: **question**, **code**, **result** (if necessary). So, the file structure could looks like:

```
--ComputerSupportSTU
  |--Task1.qmd
  |--Task1.html
  |--Task2.qmd
  |--Task2.html
```
Use a template for `qmd` document (file `template.qmd`) to generate correct html file

Example of output:

1.  Print rownames of dataset `mtcars`.

```{r}
row.names(mtcars)
```

## Task 0

1.  Create Github accout
2.  Create new repository "ComputerSupportSTU"
3.  In readme.md file add title a short repository description

## Task 1. Data Types

1. Create variables with all atomic types. Atomic types: `character`, `numeric`, `integer`, `complex`, `logical`.

2. Create the next vectors:
  * sequence from 5 to 75;
  * numbers: 3.14, 2.71, 0, 13;
  * with 100 `TRUE` values.

3. Create the next matrix:

|  |   |  |
|:---:|:-----:|:---:|
| 12  | 34    | 87  |
| -12 | -12.1 | 0   |
| 3.6 | 0.5   | 1.3 |

4. Create a `list` with all atomic types.

5. Create a `factor` with 3 levels: `infant`, `child`, `adult`.

6. Create a `data frame`

7. Change column names of the `data frame`.


## Task 2. Subsetting data. Working with NA

1. Create the next random vector:

```{r message=FALSE}
v <- rnorm(50)
```

For the vector print:

- 10th item,
- items from 10th to 20th (including 20th item),
- 10 items starting from 5th,
- all items bigger than 0,
- last 5 items.

2. Create the next data frame:

```{r message=FALSE}
df <- 
  data.frame(a = rnorm(50), b = 1:50, 
             cc = sample(letters, 50, replace = TRUE))

```

For the data frame find and print:

- the last 10 rows,
- 15th item of columns `cc`,
- subset columns `a` and `b` and rows from 10th to 20th (including 20th),
- a new vector from column `b`.

3. Create the next random vector:

```{r}
v_na <- rnorm(30)
num_na <- sample(10, 1)
v_na[sample(30, num_na)] <- NA
```

For this vector:

- print all not NA items,
- find how many NA values the vector has,
- find `min` `max` and `mean` for the vector.

## Task 3. Exploring a dataset

For this task you have to use dataset from file `airquality.csv`. The description of the data you can find by the [link](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality).

Example, how to work with the dataset. For instanse, we want to find month (column `Month`) with highest mean temperature (`Temp`)

```{r message=FALSE}
library(dplyr)

airquality |> 
  group_by(Month) |> 
  summarise(mean_temp = mean(Temp)) |> 
  filter(mean_temp == max(mean_temp))
```

You task is to answer the following questions about this data by writing appropriate code.

1. What are the column names of the data frame?

2. What are the row names of the data frame?

3. Extract the first 6 rows of the data frame and print them to the console

4. How many observations (i.e. rows) are in this data frame?

5. Extract the last 6 rows of the data frame and print them to the console

6. How many missing values are in the "Ozone" column of this data frame?

7. What is the mean of the "Ozone" column in this dataset? Exclude missing values (coded as NA) from this calculation.

8. Extract the subset of rows of the data frame where Ozone values are above 31 and Temp values are above 90.
